---
title: "Systems Serology Machine Learning Workshop - Part 1"
output:
  html_document:
    df_print: paged
---

## Getting started

Install the following packages before getting started (before the workshop):

* readxl: data import
* pheatmap: visualization
* ggplot2: visualization
* ggpubr: statistical test
* DMwR: data imputation
* ropls: multivariate analysis
* glmnet: feature selection
* RColorBrewer: color schemes for visualization

## Data import and inspection

The data is taken from the following publication: Lu et al., IFN-γ-independent immune markers of Mycobacterium tuberculosis exposure, Nature Medicine (2019).

1) First, we import the data using the function 'read_excel'. 
Note: To learn more about individual functions and their usage, use the command ?function_name, e.g., ?read_excel.

2) To vizualize the data we plot a heatmap using 'pheatmap'. 

3) We need to define X (the measurements) and y (the output we want to predict) for the further analysis.
The column names (colnames) and row names (rownames) of X should be the features and sample IDs, respectively. 

4) There are missing values in the dataset. We impute missing values with k nearest neighbors imputation using the package 'DMwR'. It is very important to evaluate how the imputation affects the model, however, this is beyond scope of this workshop.

5) The measurements are on different scales, to ensure comparison across different features we will standardize the data (Z-score).


```{r data import, fig.width = 20}
library(readxl)
library(pheatmap)
library(DMwR)

# Import the data from the excel sheet
data = read_excel("RSTR_data_NatureMedicine.xlsx")
head(data)

# Define X and y
X <- as.matrix(data[, 5:dim(data)[2]])  #Column 1:4 contains demographics and outcome 
rownames(X) <- paste0("s", seq(nrow(X))) #generates sample ids as 's1', 's2', 's3', ...
y <- factor(data$dx)
names(y) <- rownames(X)

# We make a heatmap of the raw data 
pheatmap(X)

# Data imputation for the missing vlaues. 
X <- knnImputation(X)
y <- y[match(rownames(X), names(y))] #ensures that the order of samples in Y is same as in X

# Z-score the data to bring the measurements to the same scale (mean 0 and standard deviation 1 across samples)
X_zscore <- scale(X, center = TRUE, scale = TRUE)

# Display heatmap of the z-scored data indicating phenotypic outcome
annotation_row = data.frame(phenotype = y)
rownames(annotation_row) <- rownames(X)

pheatmap(X_zscore, annotation_row = annotation_row)
```

## Description of the data

Response variable:
The phenotype we want to predict is latent Mtb infection (LTB) or resistors (RSTR). The classification is based on the diagnosis. 
Specifically, LTB and RSTR are defined in paper as following:
* LTB: A population is diagnosed based on evidence of immune sensitization to Mycobacterium tuberculosis (Mtb) antigens and the absence of clinical symptoms of tuberculosis (TB) or direct microbiologic evidence of disease.
* RSTR: a population of house hold contacts which were highly exposed to Mtb yet remained persistently negative in interferon-r (IFN-r) release assay (IGRA) and tuberculin skin tests (TST) over an average of 9.5 years of follow-up for each individual.

Features:
Different assays/tests are performed together with different antigens. The nomenclature is as follows: 
[antigen abbreviation].[test abbreviation].[metric]

List of antigen abbreviations:
ppd - purified protein derivative, proteins extracted from Mtb used in diagnosis
ag85 - antigen 85, Mtb protein
esat6 and cfp10 - Mtb protein
lam - lipoarabinomannan from Mtb cell-wall
flu - Influenza virus peptides (respiratory pathogen)
pneumo - S. pneumoniae peptides (respiratory pathogen)
cmv - cytomegalo virus (Epstein Barr)  (non-respiratory pathogen)
vzv - varicellazostervirus (non-respiratory pathogen)
rubella - rubella virus (non-respiratory pathogen)
tetanus - tetanus virus (non-respiratory pathogen)
groes - chaperone protein GroES
hspx - α-crystalline
cl - cardiolipin
ps - phosphatidylserine
b2gp1 - recombinant human apolipoprotein H/b2GP1
gp140 - HIV glycoprotein gp140
rsv - Respiratory syncytial virus

List of test abbreviations:
CD107a - Functional test of NK cell degranulation (CD107a)
IFNg - Functional test of NK cell degranulation (IFNg)
MIP1b - Functional test of NK cell degranulation (MIP1b)
ADNP - Functional test of Ab-dependent neutrophil phagocytosis 
ADCP - Functional test of Ab-dependent cellular phagocytosis
IgG.elisa - IgG ELISA	
IgG, IgM, IgG1, IgG2, IgG3, IgG4, IgA1, IgA2 - luminex FC subclasses and isotypes 
IgG1_2 -  ratio IgG1:IgG2 
IgG1_3 -  ratio IgG1:IgG3 
IgG1_4 -  ratio IgG1:IgG4 
IgG2_3 -  ratio IgG2:IgG3 
IgG3_4 -  ratio IgG3:IgG4 
FcR.3AF - luminex FcR allele variant (3AF)
FcR.3AV - luminex FcR allele variant (3AV)
FcR.2AH - luminex FcR allele variant (2AH)
FcR.2AR - luminex FcR allele variant (3AR)
G0, G0*, G0B, G0F, G0FB,... - IgG glycans

List of metric abbreviations:
AUC - area under the curve calculated from median fluorescence intensities (MFIs) from three dilutions
5, 10, 100, 1000 - dilutions

Meta data:
The dataset also includes metadata about the subjects including age, sex, and BMI that we do not use in the analysis.

## Univariate analysis

In the following, we will visualize the IgG1 values for the individual classes across antigens. We annotate the significance using the Wilcox test with the library 'ggpubr'. Note: here, the p-values are not corrected for multiple hypothesis testing!

```{r univariate}
library(ggplot2)
library(ggpubr)

# Extract indices for features including the IgG AUCs
indFeatures <- which(grepl("IgG1.1000", colnames(X_zscore))) 

# Reformat the data for the following plotting procedure
dfBox <- reshape2::melt(cbind(as.data.frame(X_zscore[,indFeatures]), class = y), id.var = "class")
colnames(dfBox) <- c("class", "features", "value")
  
ggplot(dfBox, aes(x = features, y = value, fill = class)) + 
  stat_boxplot(position = position_dodge(width = 0.75), 
               geom = 'errorbar', width = 0.2) + # boxplot 
  geom_boxplot(outlier.shape = NA) + # outliers are plotted below
  geom_point(position = position_jitterdodge(), shape = 21, size = 1, 
             aes(fill = class), color = "black") + # add points for individual measurements
  ylab("z-scored value") + 
  theme_classic() + # visualization style
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) + # rotate x-labels 
  stat_compare_means(aes(group = class), label = "p.signif", method = "wilcox.test") # comparison of groups using Wilcox test
```

## Multivariate analysis

Coming soon!


